---
title: "Hypoxia Details"
author: "Giorgi Chighladze"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Hypoxia Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Data do not always come in a format that you can readily use for analysis. Some cleaning and restructuring of data is often needed despite of the size of the dataset.

## Data 

This [data](https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-09-18) was downloaded 15 Apr 2019 from [TidyTuesday](https://github.com/rfordatascience/tidytuesday).

NEED SOME DESCRIPTION OF THE DATA and PROBLEMS IT HAS...

Before looking at the data, make sure that we do not missinterpret any entry when reading the file. For csv and txt files it is safe to read each columns as text. Also, it is good practice to read only the first few rows (5 in this case) of data for this purpose. 

```{r, warning=FALSE,comment=FALSE, message=FALSE}
library(tidyverse)
datfile <- system.file("messydata", "hypoxia.csv", package="untidydata2")
read_csv(datfile, col_types = cols(.default = "c"), n_max = 5) %>%
  knitr::kable()
```

## Problems

There are several problems that one can note at first glance. 

- variable description is included in the first row 
- column names contain spaces and symbols 
- first column is redandant and inaccurate (first row is assigned 1, even though it does not represent an observation)
- variable 'Altitude' contains two values in different units and non-numeric record ('Sea level')
- last three variables are the same as previous 3 columns, but at different oxygen content (percentge)


## Tidying Data

First we need to fix the variable names. The best way is to manually define each variable name after reading the data without column names. Since the last 6 columns represent the same variable measured at different oxygen levels, we will incorporate this piece of info in the variable names (and use later when reshaping the data).

```{r}
data <- read_csv(datfile, 
                 col_types = cols(.default = "?"), 
                 col_names = FALSE,
                 skip = 2) %>% 
  # remove first cloumn 
  select(-1) %>% 
  # set names of variables 
  set_names(c('altitude', 'air_pressure', 'ppO2', 
              'alveoli_pO2-21', 'sturation_O2-21', 'alveoli_pCO2-21',
              'alveoli_pO2-100', 'sturation_O2-100', 'alveoli_pCO2-100'))
```

After reading the data we will address remaining problems listed as follow: 

```{r}
tidy_data <- data %>%
  # set 'Sea level' as 0 ft and 0 m in format matching the other records in altitude
  mutate(altitude = ifelse(altitude == 'Sea level', '0k/0k', altitude)) %>%
  # split altitude data into columns corresponding to each unit of measurement
  separate(altitude, into = c('altitude_ft', 'altitude_m'), sep = '/') %>%
  # convert altitude into numeric data by taking into account k
  mutate_at(vars(starts_with('altitude')), parse_number) %>%
  mutate_at(vars(starts_with('altitude')), funs(. * 1000)) %>%
  # handle last 6 columns
  gather(key, value, 5:10) %>%
  # extract oxygen level (that was incorporated into the variable name earlier)
  separate(key, into = c('variable', 'O2'), sep = '-', convert = TRUE) %>%
  spread(variable, value) %>%
  arrange(O2, altitude_ft)

knitr::kable(tidy_data)
```



## Figures

Lets plot the data. 

```{r, fig.show = 'hold', fig.width = 5 }
tidy_data %>%
  mutate(O2 = factor(O2, levels = c(21, 100), labels = c('21%', '100%'))) %>%
  ggplot(aes(x = alveoli_pO2, y = altitude_m)) +
  geom_point(aes(colour = O2), na.rm = TRUE) +
  theme_light()
```


