---
title: "Tidying Grand Slam Data"
author: "Anne Eaton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A lot of the code here was originally written by Thomas Mock for the Tidy Tuesday activity on April 9, 2019. The code can be found [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-04-09/tennis_pros.rmd). To see the code used to scrape the data from Wikipedia and get it into R, see the Appendix.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The data
We will start from the dataset `grandslam_messy`, which is a messy dataset with several issues. 
```{r}
head(grandslam_messy)
```

## The issues
The main issues we want to address are 
1. We have one row for each year, one column for each tournament, and the name of the winner in each cell. Depending on the method we are going to use, we would want either one row for each unique tournament/year combination or one row for each tennis player. 

2. Sometimes two data items are in the same cell, for example, `Maud Watson (1/2)` indicates that Maud Watson won 1884 Wimbleton, but also this was her first win out of two lifetime wins. Or, in line 85, all of this appears in one cell: `Open era tennis begins  Nancy Richey (2/2)`  

3. Sometimes a cell contains a string that is not a winner's name, e.g. `tournament not created` or `World WarI`

## Tidying the data
First, ...
```{r}
dat<-grandslam_raw


# As a last step, make a data dictionary.
# This code creates a shell of documentation for the grandslam_messy dataset
# promptData(grandslam_messy, file = 'man/')
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

## Make a data dictionary for my data
 Run `?grandslam_messy` to see the data dictionary. 
 
## Appendix
The raw data was downloaded on April 15, 2019 from Wikipedia using the code below. 
```{r, eval= F}
# Pull data from wikipedia and save it into the messydata folder
library(textreadr)
library(rvest)
library(janitor)

grandslam_messy<-
  read_html("https://en.wikipedia.org/wiki/List_of_Grand_Slam_women%27s_singles_champions") %>%
  html_table(fill = TRUE) %>% 
  .[[3]] %>% 
  janitor::clean_names()  

write.table(grandslam_messy, file = 'grandslam_messy.txt')
```

