---
title: "Cost of publishing in journals"
author: "Kevin Wright"
date: "4/15/2019"
output: html_document
vignette: |
  % \VignetteIndexEntry{Cost of publishing in journals} 
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This data was downloaded 15 Apr 2019 from https://figshare.com/articles/Wellcome_Trust_APC_spend_2012_13_data_file/963054.

Inspiration from http://www.opiniomics.org/biologists-this-is-why-bioinformaticians-hate-you/
                          
```{r}
library(readxl)
datfile <- system.file("messydata", "University returns_for_figshare_FINAL.xlsx", package="untidydata2")
dat <- read_excel(datfile)
```

First thing I always do when exploring data is change the variable names to be nice. No spaces, all lowercase, short, distinct. This makes typing more pleasant while exploring the data.

```{r}
names(dat)
names(dat) <- c("id","publisher","journal","article","cost")
head(dat)
```

The first column contains a weird collection of identification data.

```{r}
head(dat$id, 30)
```

There are backslashes. Sometimes a colon is used. Sometimes a space is used. There are either 0, 1, or 2 IDs. 
